---
title: "Dates and Times Exercises"
author: "Yasmeen Aldossary"
date: "4/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library("tidyverse")
library("lubridate")
library("nycflights13")
```

## 16.2.4 Exercises
1- What happens if you parse a string that contains invalid dates?
```{r}
ymd(c("2010-10-10", "bananas"))
```
It will return NA.

2-What does the `tzone` argument to `today()` do? Why is it important?

Since different time-zones can have different dates, It determines the time-zone of the date,the value of `today()` can vary depending on the time-zone specified.


3-Use the appropriate `lubridate` function to parse each of the following dates:

```{r}
d1 <- "January 1, 2010"
d2 <- "2015-Mar-07"
d3 <- "06-Jun-2017"
d4 <- c("August 19 (2015)", "July 1 (2015)")
d5 <- "12/30/14" # Dec 30, 2014
```

```{r}
mdy(d1)
ymd(d2)
dmy(d3)
mdy(d4)
mdy(d5)
```


## 16.3.4 Exercises

1- How does the distribution of flight times within a day change over the course of the year?

```{r}
flights_dt %>%
  mutate(date = make_date(year(dep_time),
                          month(dep_time),
                          mday(dep_time)),
         hour = hour(dep_time)) %>%
  group_by(date, hour) %>%
  ggplot(mapping = aes(x = hour, group = date)) +
  geom_density(alpha = .1)
```
The distributions for most days are very much the same.

2- Compare `dep_time`, `sched_dep_time` and `dep_delay`. Are they consistent? Explain your findings.

Check if there are any inconsistent records:
```{r}
flights_dt %>% select(contains('dep')) %>%
  mutate(cal_delay = as.numeric(dep_time - sched_dep_time) / 60) %>%
  filter(dep_delay != cal_delay)
```

These inconsistent records have one thing in common. The dep_delay tells us the flights were delayed, but cal_delay tells us the flights departed early. The reason is that these delayed flights actually departed on the next day, and were not reflected in dep_time. If we add one day to the dep_time, the results should be consistent.

```{r}
flights_dt %>% select(contains('dep')) %>%
  mutate(cal_delay = as.numeric(dep_time - sched_dep_time) / 60) %>%
  filter(dep_delay != cal_delay) %>%
  mutate(dep_time = update(dep_time, mday = mday(dep_time) + 1)) %>%
  mutate(cal_delay = as.numeric(dep_time - sched_dep_time)) %>%
  filter(dep_delay != cal_delay)
```
All records are now consistent.


3- Compare air_time with the duration between the departure and arrival.

```{r}
flights_dt %>%
  mutate(
    flight_duration = as.numeric(arr_time - dep_time),
    air_time_mins = air_time,
    diff = flight_duration - air_time_mins
  ) %>%
  select(origin, dest, flight_duration, air_time_mins, diff)
```


4- How does the average delay time change over the course of a day? Should you use dep_time or sched_dep_time? Why?

Use sched_dep_time because that is the relevant metric for someone scheduling a flight. Also, using dep_time will always bias delays to later in the day since delays will push flights later.

```{r}
flights_dt %>%
  mutate(sched_dep_hour = hour(sched_dep_time)) %>%
  group_by(sched_dep_hour) %>%
  summarise(dep_delay = mean(dep_delay)) %>%
  ggplot(aes(y = dep_delay, x = sched_dep_hour)) +
  geom_point() +
  geom_smooth()
```

5- On what day of the week should you leave if you want to minimise the chance of a delay?

```{r}
flights_dt %>%
  mutate(dayweek = wday(sched_dep_time, label = TRUE)) %>%
  group_by(dayweek) %>%
  summarize(avg_dep_delay = mean(dep_delay, na.rm = TRUE),
            avg_arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  gather(key = 'delay', value = 'minutes', 2:3) %>%
  ggplot() +
  geom_col(mapping = aes(x = dayweek, y = minutes, fill = delay),
           position = 'dodge')
```

It looks like Saturdays are the best days to fly.

6- What makes the distribution of diamonds$carat and flights$sched_dep_time similar?

```{r}
ggplot(diamonds, aes(x = carat)) +
  geom_density()
```

In both carat and sched_dep_time there are abnormally large numbers of values are at nice “human” numbers. In sched_dep_time it is at 00 and 30 minutes. In carats, it is at 0, 1/3, 1/2, 2/3,

```{r}
ggplot(diamonds, aes(x = carat %% 1 * 100)) +
  geom_histogram(binwidth = 1)
```

In scheduled departure times it is 00 and 30 minutes, and minutes ending in 0 and 5.
```{r}
ggplot(flights_dt, aes(x = minute(sched_dep_time))) +
  geom_histogram(binwidth = 1)
```


7- Confirm my hypothesis that the early departures of flights in minutes 20-30 and 50-60 are caused by scheduled flights that leave early. Hint: create a binary variable that tells you whether or not a flight was delayed.

First, I create a binary variable early that is equal to 1 if a flight leaves early, and 0 if it does not. Then, I group flights by the minute of departure. This shows that the proportion of flights that are early departures is highest between minutes 20–30 and 50–60.

```{r}
flights_dt %>% 
  mutate(minute = minute(dep_time),
         early = dep_delay < 0) %>% 
  group_by(minute) %>% 
  summarise(
    early = mean(early, na.rm = TRUE),
    n = n()) %>% 
  ggplot(aes(minute, early)) +
    geom_line()
```


## 16.4.5 Exercises

1- Why is there months() but no dmonths()?

Months do not have a fixed duration in seconds, unlike days, weeks, and years, becauses ther are months with 28, 29, 30, and 31 days.

2- Explain days(overnight * 1) to someone who has just started learning R. How does it work?

`overnight` is a boolean variable. If arr_time is less than dep_time, then the flight arrives on the next day, and overnight is TRUE; otherise, FALSE. Actually, the underlying value of TRUE is 1, and FALSE is 0, so the * 0 can actually be omitted.

If overnight is TRUE, or 1, days(overnight) will add one day to the arr_time and sched_arr_time datetime.

```{r}
flights_dt %>% 
  mutate(
    overnight = arr_time < dep_time,
    arr_time = arr_time + days(overnight),
    sched_arr_time = sched_arr_time + days(overnight)
  )
```
  
3- Create a vector of dates giving the first day of every month in 2015. Create a vector of dates giving the first day of every month in the current year.
```{r}
year_2015 <- years(2015) + months(c(1:12)) + days(1)
year_2015
```

```{r}
year_current <- years(year(today())) + months(c(1:12)) + days(1)
year_current
```

4- Write a function that given your birthday (as a date), returns how old you are in years.

```{r}
age <- function(bday) {
  (bday %--% today()) %/% years(1)
}
age(ymd("1990-10-12"))
```

5- Why can’t `(today() %--% (today() + years(1))) / months(1)` work?

```{r}
(today() %--% (today() + years(1))) / months(1)
```
The interval is defined as starting from today to a year from today, and there 12 months in between.

