---
title: "Relational Data Exercises"
author: "Yasmeen Aldossary"
date: "4/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(nycflights13)
```



## 13.2.1 Exercises
1- Imagine you wanted to draw (approximately) the route each plane flies from its origin to its destination. What variables would you need? What tables would you need to combine?

To draw the route each plane flies from its origin to its destination it requires to joins two tables the flights and airports tables. we need from the flights table two variables the `origin` and destination `dest`  airport of each flight. And from the airports table the longitude `lon` and latitude `lat` of each airport.

```{r}
flights_latlon <- flights %>%
  inner_join(select(airports, origin = faa, origin_lat = lat, origin_lon = lon),
    by = "origin"
  ) %>%
  inner_join(select(airports, dest = faa, dest_lat = lat, dest_lon = lon),
    by = "dest"
  )
```

2- I forgot to draw the relationship between weather and airports. What is the relationship and how should it appear in the diagram?

```{r}
colnames(weather)
```

```{r}
colnames(airports)
```
The two tables can be matched vie the variable `origin` in `weather` and `faa` in `airports`.

3- weather only contains information for the origin (NYC) airports. If it contained weather records for all airports in the USA, what additional relation would it define with flights?

If the weather was included for all airports in the US, then it would provide information about the weather at the destination airport at the time of the flight take off, unless the arrival date-time were calculated.


4- We know that some days of the year are “special”, and fewer people than usual fly on them. How might you represent that data as a data frame? What would be the primary keys of that table? How would it connect to the existing tables?

We can create a new table `special` containing the pertaining information of the special dates. The primary key of the table would be the (year, month, day) columns. The (year, month, day) columns could be used to join `special` with other tables.



## 13.3.1 Exercises

1- Add a surrogate key to flights.
```{r}
flights %>%
  mutate(flight_id = row_number()) %>%
  select(flight_id, everything())
```


2- Identify the keys in the following datasets:
1. `Lahman::Batting`
2. `babynames::babynames`
3. `nasaweather::atmos`
4. `fueleconomy::vehicles`
5. `ggplot2::diamonds`

The primary keys for `Lahman::Batting` are `playerID`, `yearID`, `stint`, `teamID`, and `lgID`.

```{r}
Lahman::Batting %>%
  count(playerID, yearID, stint) %>%
  filter(n > 1) %>%
  nrow()
```

The primary keys for `babynames::babynames` are `year`, `sex`, and `name`.

```{r}
babynames::babynames %>%
  count(year, sex, name) %>%
  filter(n > 1) %>%
  nrow()
```

The primary keys for nasaweather::atoms are lat, long, year, and month

```{r}
nasaweather::atmos %>%
  count(lat, long, year, month) %>%
  filter(n > 1) %>%
  nrow()
```
The column id, the unique EPA identifier of the vehicle, is the primary key for fueleconomy::vehicles.

```{r}
fueleconomy::vehicles %>%
  count(id) %>%
  filter(n > 1) %>%
  nrow()
```

There is no primary key for ggplot2::diamonds since there is no combination of variables that uniquely identifies each observation.
```{r}
ggplot2::diamonds %>%
  distinct() %>%
  nrow()

nrow(ggplot2::diamonds)

```


3- Draw a diagram illustrating the connections between the Batting, People, and Salaries tables in the Lahman package. Draw another diagram that shows the relationship between People, Managers, AwardsManagers.
How would you characterise the relationship between the Batting, Pitching, and Fielding tables?

The variables names are printed below:
```{r}
colnames(Lahman::Batting)
```

```{r}
colnames(Lahman::Master)
```

```{r}
colnames(Lahman::Salaries)
```

```{r}
colnames(Lahman::Managers)
```

```{r}
colnames(Lahman::AwardsManagers)
```

`Batting` contains batting statistics for players. The primary keys are `playerID`, `yearID`, `stint`, `teamID`, and `lgID`. The players’ biographical information are stored in Master and can be matched with playerID. The salary information for each player in each year can be matched with `playerID`, `yearID`, `teamID`, and `lgID`.

`Batting` and `Managers` can be matched with `playerID`, `yearID`, `teamID`, and `lgID`. `Mangers` and `AwardManagers` are matched with `playerID`, `yearID`, `teamID`, and `lgID`.

`Batting`, `Pitching`, and `Fielding` can be matched with `playerID`, `yearID`, `stint`, `teamID`, and `lgID`.



## 13.4.6 Exercises

1- Compute the average delay by destination, then join on the airports data frame so you can show the spatial distribution of delays. Here’s an easy way to draw a map of the United States:

```{r}
airports %>%
  semi_join(flights, c("faa" = "dest")) %>%
  ggplot(aes(lon, lat)) +
    borders("state") +
    geom_point() +
    coord_quickmap()
```

You might want to use the size or colour of the points to display the average delay for each airport.

```{r}
avg_dest_delays <-
  flights %>%
  group_by(dest) %>%
  summarise(delay = mean(arr_delay, na.rm = TRUE)) %>%
  inner_join(airports, by = c(dest = "faa"))

avg_dest_delays %>%
  ggplot(aes(lon, lat, colour = delay)) +
  borders("state") +
  geom_point() +
  coord_quickmap()
```

2- Add the location of the origin and destination (i.e. the lat and lon) to flights.
To better distingush the added location information between origin and destination, we can define suffix to the names of the variables joined to the exisiting table. To illustrate:

```{r}
flights %>% 
  left_join(airports, by = c('dest' = 'faa')) %>%
  left_join(airports, by = c('origin' = 'faa'), suffix = c('.dest', '.origin')) %>%
  select(dest, origin, contains('lat'), contains('lon'))
```

3- Is there a relationship between the age of a plane and its delays?

We merge `flights` with the `planes`, To compare the age of the plane to flights delay, which contains a variable `plane_year`, with the year in which the plane was built. To look at the relationship between plane age and departure delay, we'll calculate the average departure delay for each age of a flight.

```{r}
plane_cohorts <- inner_join(flights,
  select(planes, tailnum, plane_year = year),
  by = "tailnum"
) %>%
  mutate(age = year - plane_year) %>%
  filter(!is.na(age)) %>%
  mutate(age = if_else(age > 25, 25L, age)) %>%
  group_by(age) %>%
  summarise(
    dep_delay_mean = mean(dep_delay, na.rm = TRUE),
    dep_delay_sd = sd(dep_delay, na.rm = TRUE),
    n_dep_delay = sum(!is.na(dep_delay))
  )

ggplot(plane_cohorts, aes(x = age, y = dep_delay_mean)) +
  geom_point() +
  scale_x_continuous("Age of plane (years)", breaks = seq(0, 30, by = 10)) +
  scale_y_continuous("Mean Departure Delay (minutes)")

```
The average departure delay is increasing for planes with ages up until 10 years. After that the departure delay decreases or levels off.


4- What weather conditions make it more likely to see a delay?

```{r}
flight_weather <-
  flights %>%
  inner_join(weather, by = c(
    "origin" = "origin",
    "year" = "year",
    "month" = "month",
    "day" = "day",
    "hour" = "hour"
  ))

flight_weather %>%
  group_by(precip) %>%
  summarise(delay = mean(dep_delay, na.rm = TRUE)) %>%
  ggplot(aes(x = precip, y = delay)) +
  geom_line() + geom_point()

flight_weather %>%
  ungroup() %>%
  mutate(visib_cat = cut_interval(visib, n = 10)) %>%
  group_by(visib_cat) %>%
  summarise(dep_delay = mean(dep_delay, na.rm = TRUE)) %>%
  ggplot(aes(x = visib_cat, y = dep_delay)) +
  geom_point()


```

Almost any amount of precipitation is associated with a delay. However, there is not a strong a trend above 0.02 in. of precipitation. There seems to be a stronger relationship between visibility and delay. Delays are higher when visibility is less than 2 miles.

5-What happened on June 13 2013? Display the spatial pattern of delays, and then use Google to cross-reference with the weather.

There was a large series of storms (derechos) in the southeastern US. The following plot show that the largest delays were in Tennessee (Nashville), the Southeast, and the Midwest, which were the locations of the derechos.

```{r}
flights %>% filter(year == 2013, month == 6, day == 13) %>%
  group_by(dest) %>%
  summarize(avg_arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  left_join(airports, by = c('dest' = 'faa')) %>%
  ggplot(aes(x = lon, y = lat, size = avg_arr_delay, color = avg_arr_delay)) +
  borders('state') +
  geom_point(alpha = .5) +
  scale_color_continuous(low = 'yellow', high = 'red') + 
  coord_quickmap()
```


## 13.5.1 Exercises

1- What does it mean for a flight to have a missing tailnum? What do the tail numbers that don’t have a matching record in planes have in common? (Hint: one variable explains ~90% of the problems.)

Flights have a missing tailnum are those that were cancellled, or without missing dep_time, etc.

```{r}
flights %>%
  filter(is.na(tailnum))
```

For those tailnum that don’t have a matching record in plane, it seems most of them come from the same two carriers.

```{r}
flights %>%
  anti_join(planes, by = 'tailnum') %>%
  group_by(carrier) %>%
  summarize(n = n()) %>%
  arrange(desc(n))
```


2- Filter flights to only show flights with planes that have flown at least 100 flights.

```{r}
flights %>%
  filter(!is.na(tailnum)) %>%
  group_by(tailnum) %>%
  mutate(n = n()) %>%
  filter(n >= 100)
```

3- Combine fueleconomy::vehicles and fueleconomy::common to find only the records for the most common models.

```{r}
fueleconomy::vehicles %>%
  semi_join(fueleconomy::common, by = c("make", "model"))

fueleconomy::vehicles %>%
  distinct(model, make) %>%
  group_by(model) %>%
  filter(n() > 1) %>%
  arrange(model)

fueleconomy::common %>%
  distinct(model, make) %>%
  group_by(model) %>%
  filter(n() > 1) %>%
  arrange(model)
```

4- Find the 48 hours (over the course of the whole year) that have the worst delays. Cross-reference it with the weather data. Can you see any patterns?

```{r}
delay <- flights %>%
    filter(dep_delay > 0, arr_delay > 0) %>%
    group_by(month, day) %>%
    summarise(count = n(),
              mean_delay = mean(dep_delay+arr_delay, na.rm =T)) %>%
    mutate(average_twoday = mean_delay + lag(mean_delay)) %>%
    arrange(desc(average_twoday))
kable(head(delay))
```

```{r}
twoday_delay <- delay %>%
    inner_join(weather, by = c("month", "day"))
twoday_delay %>%
    ggplot() +
    geom_point(mapping = aes(x = humid, y = average_twoday )) +
    labs(y = 'Average Delay', x = 'Humidity')
```
It's hard to see any patterns in weather when we calculate for such wide range of time (48 hours) because weather usually change in a smaller window of time (when it rains, it usually last a few hours).


5- What does anti_join(flights, airports, by = c("dest" = "faa")) tell you? What does anti_join(airports, flights, by = c("faa" = "dest")) tell you?

```{r}
anti_join(flights, airports, by = c("dest" = "faa"))

```

Drops all observations in flights that have a match in airport.

```{r}
anti_join(airports, flights, by = c("faa" = "dest"))
```

Drops all observations in airport that have a match in flights.

6- You might expect that there’s an implicit relationship between plane and airline, because each plane is flown by a single airline. Confirm or reject this hypothesis using the tools you’ve learned above.


First, find all distinct airline, plane combinations.
```{r}
planes_carriers <-
  flights %>%
  filter(!is.na(tailnum)) %>%
  distinct(tailnum, carrier)

```
The number of planes that have flown for more than one airline are those `tailnum` that appear more than once in the `planes_carriers` data.

```{r}
planes_carriers %>%
  count(tailnum) %>%
  filter(n > 1) %>%
  nrow()
```

The names of airlines are easier to understand than the two-letter carrier codes. The airlines data frame contains the names of the airlines.

```{r}
carrier_transfer_tbl <- planes_carriers %>%
  # keep only planes which have flown for more than one airline
  group_by(tailnum) %>%
  filter(n() > 1) %>%
  # join with airlines to get airline names
  left_join(airlines, by = "carrier") %>%
  arrange(tailnum, carrier)

carrier_transfer_tbl
```
